Coupling and uncoupling of midline morphogenesis and cell flow in amniote gastrulation

Version 1

Abstract

Summary

Large-scale cell flow characterizes gastrulation in animal development. In amniote gastrulation, a bilateral vortex-like counter-rotating cell flow, called ‘polonaise movements’, appears along the midline. Here, through experimental manipulations, we addressed relationships between the polonaise movements and morphogenesis of the primitive streak, the earliest midline structure in amniotes. Suppression of the Wnt/planar cell polarity (PCP) signaling pathway maintains the polonaise movements along a deformed primitive streak. Mitotic arrest leads to diminished extension and development of the primitive streak and maintains the early phase of the polonaise movements. Ectopically induced Vg1, an axis-inducing morphogen, generates the polonaise movements, aligned to the induced midline, but disturbs the stereotypical cell flow pattern at the authentic midline. Despite the altered cell flow, induction and extension of the primitive streak are preserved along both authentic and induced midlines. Finally, we show that ectopic axis-inducing morphogen, Vg1, is capable of initiating the polonaise movements without concomitant PS extension under mitotic arrest conditions. These results are consistent with a model wherein primitive streak morphogenesis is required for the maintenance of the polonaise movements, but the polonaise movements are not necessarily responsible for primitive streak morphogenesis. Our data describe a previously undefined relationship between the large-scale cell flow and midline morphogenesis in gastrulation.

Introduction

Large-scale cell flow during gastrulation is an evolutionarily conserved biological phenomenon in embryogenesis (1-3). During this period, animals with bilateral symmetry, called bilaterians, initiate development of three germ layers (i.e. the ectoderm, mesoderm and endoderm) and midline structures (e.g. the notochord and primitive streak) along the midline axis (4-6). The large-scale cell flow has been linked to early embryonic morphogenesis, including axial structures, through rearrangement of cells and/or transportation of signaling molecules (2, 3). Coupling of the large-scale cell flow and body axis morphogenesis has been extensively studied in invertebrate bilaterians, particularly insect models (7-9). However, in vertebrates, our understanding of whether and how the large-scale cell flow and midline morphogenesis are coupled is limited.

Vertebrates are classified into two groups: non-amniotes (e.g. fish and amphibians) and amniotes (e.g. birds and mammals) (1). In non-amniotes, particularly amphibians, the large-scale cell flow is coupled with formation of the notochord through Wnt/planar cell polarity (PCP) pathway-regulated convergent extension, which does not depend on mitosis (10, 11). In amniotes, a bilateral vortex-like counter-rotating cell flow, termed ‘polonaise movements’, occurs within the epiblast along the midline axis, prior to and during primitive streak (PS) formation (12-14). The PS is an evolutionarily unique embryonic structure in amniotes, which is not only an organizing center for gastrulation but also the earliest midline structure (6, 15-17). During amniote gastrulation, widespread mitosis has been identified throughout the embryo, including the PS (18), and previous studies showed that mitotic arrest leads to morphologically diminishing PS extension (19, 20). Further, the Wnt/PCP pathway also plays an important role for proper PS extension and patterning (21). Since the chick embryo is readily accessible and closely represents the human gastrula (22), much of what is known about the cell flow and PS development has come from studies using this model system (23). The PS is first visible as a cell cluster at the posterior embryonic disc at the initial streak stage [HH2; Hamburger and Hamilton staging (24)] formed through the influence of several axis inducing factors [e.g. Vg1 and Chordin (25, 26)]; it then extends anteriorly along the midline axis (6, 15, 16). During PS development, displacement of cells and/or tissue tension potentially generates biomechanical forces, which lead to cell flows and/or PS morphogenesis (20, 21, 27, 28). Recent studies discuss that cell intercalation and ingression underlying PS formation, leading to the polonaise movements through altering the neighbor tissue tension (21, 28, 29), suggesting that the polonaise movements are likely a consequence of PS morphogenesis (16, 30, 31). Other studies demonstrate that the polonaise movements, generated by the myosin-cable-mediated tissue tension with cell intercalation and ingression, lead to PS morphogenesis (20, 27, 31). Thus, the relationship between the polonaise movements and PS morphogenesis remains controversial (31).

Here, we characterize a previously undefined relationship between the polonaise movements and PS morphogenesis. Suppression of the Wnt/PCP pathway generates a wider and shorter PS while maintaining the polonaise movements. Mitotic arrest diminishes PS extension and maintains the early phase of the polonaise movements. Moreover, experimentally manipulating the large-scale cell flow by inducing a secondary midline axis indicates that the authentic PS forms despite altered cell flows. Ultimately, a single axis-inducing morphogen, Vg1, is capable of initiating the polonaise movements despite defective PS morphogenesis. These results suggest that PS morphogenesis is not responsible for the initiation and early phase of the polonaise movements.

Further, the polonaise movements are not necessarily responsible for PS morphogenesis. This study suggests that amniote gastrulation may be evolutionarily distinct from non-amniote gastrulation in the mechanisms coupling large-scale cell flow and midline morphogenesis.

Results

Suppression of the Wnt/PCP pathway maintains the bilateral vortex-like counter-rotating cell flow along a deformed PS

Throughout this work, the cell flows were recorded by tagging individual epiblast cells with electroporated fluorescent tracers at the pre-streak stage (32). The tagged embryos were live-imaged until HH3 with a modified New culture system, which we previously described (33, 34). To reconstruct trajectories of the tagged cells and quantitatively analyze the cell flow pattern, the recorded images were processed using tracking analysis tools (35), Particle Image Velocimetry (PIV) (36-40), and subsequent visualization techniques (41, 42).

To address the relationship between the polonaise movements and PS morphogenesis, we experimentally manipulated the shape of the PS while maintaining the population of PS cells (Fig. 1A-B). The DEP domain of Dishevelled (Dsh; a transducer protein of Wnt signaling) is responsible for the non-canonical Wnt/PCP pathway (43, 44), and misexpression of dominant-negative Dsh lacking DEP [dnDsh(ΔDEP)] leads to deformation of the midline structures, including the PS (21). Further, the Wnt/PCP pathway is involved in cellular polarity and migration, while the canonical Wnt pathway regulates cell proliferation (45). We generated a dnDsh(ΔDEP)-GFP construct, we will refer to this as ΔDEP-GFP, and tested whether the construct was capable of modifying cellular polarity, resulting in PS deformation (Fig. 1C-E, Fig. S1A-D). In the control-GFP-introduced PS and epiblast cells, Pk1 (a PCP core component) exhibited a polarized localization at the E-cadherin-based adherens junctions, particularly cell-cell junctions (Fig. S1A-B, n=3). In contrast, this Pk1-localization pattern was not identified in the ΔDEP-GFP-misexpressing PS and epiblast cells (Fig. S1C-D, n=3), implying that the construct efficiently disrupted the cellular polarity within the misexpressing cells. The expression of Brachyury indicated that ΔDEP-GFP-PS cells maintained their identity (Fig. 1E) but displayed a shorter and wider morphogenesis than the control PS (Fig. 1C-2E, n≥4 for each). The number of PS cells was not significantly different from the control- and ΔDEP-GFP-PSs (control 8.5±0.4x103 cells vs ΔDEP 8.8±1.2x103 cells, p=0.68, n=3 for each, Fig. 1F). These results indicated that ΔDEP misexpression was capable of suppressing the Wnt/PCP pathway through disrupting the localization pattern of the PCP core component while maintaining the number of the PS cells. Having confirmed the function of the ΔDEP-GFP construct, we applied it as a molecular tool for manipulating the shape of the PS.

‘Polonaise movements’ persist under suppression of Wnt/PCP pathway.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and the Wnt/PCP pathway. A-P and L-R; anterior-posterior and left-right body axes, respectively. (B) Diagram of experimental set up. (C, D) In situ hybridizations for Brachyury (Bra) in control- and ΔDEP-GFP-misexpressing embryos. (E) Overlay of embryos hybridized for Bra (n=4 for each). (F) Cell number in PS [control 8.5±0.4x103 cells vs ΔDEP 8.8±1.2x103 cells, p=0.68, n=3 for each (two-tailed Student’s t test)]. (G-H’’’) Trace of cell flow path of electroplated epiblast cells with Flowtrace. The initiation of the polonaise movements was set as Time 0 (t=0). See also Movie S1. (I-J’’’) Vector fields overlaid with fluorescent images, generated from the images of the control (G-G’’’) and aphidicolin-treated embryos (H-H’’’). (K-L’’’) Streamlines, visualizing averaged cell flows during each time period. (M-N’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period. Blue, clockwise; red, counter-clockwise rotation.

Early phase of ‘polonaise movements’ remains under global mitotic arrest.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and mitosis. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) BrdU incorporation in control-sham operated and aphidicolin treated in embryos. White boxes show enlarged areas. (C’, C”) High magnification of boxed areas in PS and non-PS, respectively. (E) In situ hybridizations for Brachyury (Bra), Vg1, and Sox3 in control and aphidicolin-treated embryos. Arrow heads indicate diminished PSs. Black boxes, enlarged area including the diminished PS. (F-G’’’) Trace of cell flow path with Flowtrace analysis (green is trace, yellow indicates endpoint) of fluorescently-tagged epiblast cells. The initiation of the polonaise movements was set as Time 0 (t=0). See also Movie S2. (H-I’’’) Vector fields, generated from the images of control (F-F’’’) and aphidicolin-treated embryos (G-G’’’). (J-K’’’) Streamlines, visualizing averaged cell flows during each time period. Interpolated vectors are displayed in orange. (L-M’’’) Vorticity plots, displaying averaged measure of the local rotation during each time period. Blue, clockwise; red, counter-clockwise rotation. Note: Since vorticity is calculated for all deviation from set point, slight curves and full rotations receive the same color indication, as seen in M’’’ which is not maintaining a bilateral vortex-like-rotating cell flows. See Materials and Methods section for a full discussion.

To address a relationship between the polonaise movements and the shape of the PS, we live-imaged the control- and ΔDEP-GFP-embryos from pre-streak stage to HH3 and analyzed the resulting cell flows (Fig. 1G-N”’, and movie S1). In the control-GFP embryo, the cell flow pattern, which was visualized by tracking tools and quantitatively analyzed by PIV with subsequent flow-visualization techniques (41, 42), identified robust polonaise movements (Fig. 1G-G”’, I-I”’, K-K”’, M-M”’, and movie S1, n=5,). Surprisingly, these analyses indicated that suppression of the Wnt/PCP pathway maintained the polonaise movements despite PS deformation (n=5, Fig. 1C-N”’ and movie S1). While the topology of the counter-rotating flow pattern were maintained under suppression of the Wnt/PCP pathway, the distance between the left-right rotating cell flows was wider in the ΔDEP embryos than the controls (control 860.2±86.2μm vs ΔDEP 1008.4±183.8μm, p=0.026, n=4 for each, Fig. S1E-I), implying that the shape of the PS and/or the Wnt/PCP signaling-mediated cell intercalation affects the location of these rotations. These data demonstrated that the polonaise movements persisted even though the PS as a midline structure was deformed under the condition of suppressing the Wnt/PCP pathway.

Mitotic arrest diminishes PS formation but initially preserves the bilateral vortex-like rotating cell flow

The above data showed that a deformed midline structure allowed the polonaise movements to occur while the position of the LR rotations in the polonaise movements was affected (Fig. 1 and Fig. S1). However, it remained unclear the relationship between the polonaise movements and the population of the PS cells, particularly expansion of the PS precursor cells in PS morphogenesis (Fig. 2A). To address this question, we applied a mitotic arrest approach using aphidicolin (a specific DNA polymerase inhibitor), which has been shown to diminished PS development (19, 20). In chick embryonic disc, active cell proliferation was identified in both the epiblast and PS cells (Fig. 2C)(18). As verified by BrdU incorporation assay, the aphidicolin-treated embryos exhibited little or no mitotic activity (Fig. 1C-D). Morphogenesis of PS was monitored by Brachyury (a marker for PS and mesodermal cells) and Sox3 (an epiblast marker). The expression domain and midline extension of cells expressing Brachyury and negative for Sox3 were considerably reduced in the aphidicolin-treated embryos, compared to the control (Fig. 1E; n≥4 for each). The expression pattern of Brachyury and PS extension followed an aphidicolin dose-dependent diminishment (Fig. S1A-D). The Vg1-expressing area was also diminished but detectable in the aphidicolin-treated embryos (Fig. 1E, n=4), suggesting that the axis-inducing morphogen was preserved in the embryonic disc, but expansion of the PS precursor cells was suppressed by mitotic arrest. These results indicate that mitotic arrest, induced by the aphidicolin-treatment, remarkably reduced PS extension and development, consistent with the previous works (19, 20).

To address the relationship between the polonaise movements and the population of the PS cells, we live-imaged the fluorescently tagged embryos under aphidicolin-treatment and analyzed the resulting cell flow (Fig. 2F-L”’ and movie S2). Tracking and PIV-based analyses visualized the cell flow pattern in control and aphidicolin-treated embryos (Fig. 2F-M”’). In control embryos, the polonaise movements occurred from pre-streak until HH3 for more than four hours (n=4, Fig. 2F-F”’ and movie S2). The streamline and vorticity plots illustrated that the polonaise movements became more robust in a time-dependent manner (Fig. 2H-H”’, J-J”’ and L-L”’). Strikingly, in the aphidicolin-treated embryos (n=4), the polonaise movements initiated but were maintained for a shorter period, 2.2±1.2 hours (control 6.5±1.2 hours, p=0.0038, Fig. 2G-G”’, movie S1). The streamline and vorticity plots also indicated the initiation and early phase of the polonaise movements in the aphidicolin-treated embryos (Fig. 2I-I”’, K-K’”, and M-M”’). Together, these data demonstrated that the typical pattern of the polonaise movements was induced and initially maintained in the short-term, while there was no or little indication of robust PS formation or extension.

Aphidicolin potentially induces apoptosis (46), which might indirectly lead to the cell flow changes through tissue tension alterations. To test this possibility, we examined apoptosis in the control and aphidicolin-treated embryos (Fig. S3A-C). TUNEL assay showed that the aphidicolin-treatment slightly increased apoptotic rate but there was no significant difference between these embryos and controls (control sham, 10.6±3.5%; aphidicolin, 12.1±3.7%, p=0.20, n=4 for each, Fig. S3A-C), suggesting that the bilateral rotating cell flow in the aphidicolin-treated embryos was not indirectly induced by apoptosis.

Previous studies have shown that a large-scale ring-like distribution of phospholyrated-myosin (pMyosin) cables among the epiblast cell layer is required for tissue tension related to cell flow (20, 27, 28). We examined whether the pMyosin cables were maintained in the treated embryos over the course of the imaging (Fig. S3D-F). In the aphidicolin-treated embryos, the large-scale ring-like distribution pattern of the pMyosin cables remained in the epiblast (Fig. S3F), similar to the control embryos (Fig. S3D), implying that the mitotic arrested-embryos maintained tissue tension, consistent with the previous report (20). Thus, these results support a model that mitosis is coupled with PS morphogenesis but is not necessarily required for the initiation and early phase of the polonaise movements.

PS properly extends along the midline despite altering the large-scale cell flow

Given that PS morphogenesis is required for maintenance of the polonaise movements, we next examined the converse of the relationship; that the polonaise movements may be required for PS formation. To test this model, we experimentally reprogrammed the large-scale cell flow (Fig. 3A-B). Ectopically induced Vg1 leads to formation of a Brachyury-positive secondary PS and a secondary midline axis (Fig. 3C-D)(47-49). COS cells expressing either a control-GFP (control-GFP/COS) or Vg1 (Vg1/COS) construct were placed at the anterior marginal zone (AMZ) at pre-streak stage, approximately 180° from the posterior marginal zone where the authentic PS forms, and we analyzed the resulting cell flow during both authentic and ectopically induced PS extension (Fig. 3B). If the bilateral rotating cell flow is required for PS formation, both the authentic and induced PSs should each be accompanied by polonaise movements during the extension. The trajectory analysis by Imaris and PIV-based analyses showed that the control-COS implanted embryos exhibited stereotypical polonaise movements along the authentic midline axis, indicating that the COS cell implantation maintained an authentic cell flow pattern (Fig. 3F, Fig. S4, and movie S3, n=4). In contrast, analyses of the cell flow in the Vg1/COS implanted embryos illustrated the disruption of the authentic polonaise movements at the extending authentic PS (Fig. 3F-F”’, H-H”’, J-J”’, L-L”’, Fig. S5, S6 and movie S3, S4, n=4). At pre-streak stage, the epiblast cells initially converged to the center of the embryonic disc along both the authentic and the induced axes, and the bilateral rotating cell flows began to appear at both axes (Fig. 3F, Fig. S5, S6, and movie S3, S4). This cell flow pattern continued for approximately one hour (Fig. 3J’, J”, Fig. S5, and movie S3). Before extension of the authentic PS, the bilateral rotating cell flow along the induced axis became dominant, and eventually overrode the original cell flow (Fig. 3J’, J”, Fig. S5 and movie S3). Despite global disruption and reversed counter-rotating cell flows, the authentic PS extended properly along the authentic midline (Fig. 3D, F”’, H”’, J”’, L”’, Fig. S5, and movie S3; n=4). Secondary midline-axis induction by ectopic Vg1 created the polonaise movements around the secondary PS and altered the original cell flow pattern, generating a reprogrammed large-scale cell flow (Fig. 3F-F”’, H-H”’, J-J”’, L-L”’G, Fig. S5, S6 and movie S3, S4). These results indicate that extension of the authentic PS is preserved even under the defective polonaise movements.

Authentic PS extends under disruption of authentic ‘polonaise movements’.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, using ectopic Vg1-inducing system. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) In situ hybridizations for Brachyury (Bra) in control and ectopic Vg1-induced PS, respectively. (E-F’’’) Trace of cell flow path of electroplated epiblast cells by using Imaris tracking analysis. See also Movie S3. White circles, COS cell implanted site. The initiation of the polonaise movements was set as Time 0 (t=0). (G-H’’’) Vector fields, generated from the images of control-GFP/COS-(E-E’’’) and Vg1/COS-implanted embryos (F-F’’’). (I-J’’’) Streamlines, visualizing averaged cell flows during each time period of the vector fields. (K-L’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period in the vector fields. Blue, clockwise; red, counter-clockwise rotation. Note: vorticity is identified in both a full-rotation and curve, particularly in (L-L’’’). The white asterisks indicate the polonaise movements at the induced PS which are opposite direction to the authentic PS. (M) Summary of the cell flow patterns in control-GFP/COS- and Vg1/COS-implanted embryos.

Vg1 is capable of inducing the bilateral vortex-like rotating cell flow despite diminished PS extension under mitotic arrest

The above data indicate the lack of a strong relationship between the polonaise movements and PS morphogenesis; however, it remained unclear how the polonaise movements were induced. Given that mitotic arrest maintained both Vg1-expression and the initiation of the polonaise movements (Fig. 2 and movie S2) and that induction of the secondary midline axis by ectopic Vg1 reprograms the epiblast cell flow (Fig. 3, Fig. S4, S5, and movie S3), we tested the possibility that the axis-inducing morphogen, Vg1, would be capable of inducing the polonaise movements in the absence of PS extension (Fig. 4A). We combined the secondary axis induction by AMZ implantation of Vg1/COS cells with aphidicolin-treatment and analyzed the resulting cell flow (Fig. 4B, Fig. S6, S7, and movie S4). The BrdU assay showed that the aphidicolin treatment effectively blocked mitosis in Vg1/COS implanted embryos (control sham 88±9.0% vs control-GFP/COS with aphidicolin 0.2±0.1% vs Vg1/COS with aphidicolin 0.19±0.2%, p= 0.0000005, n=3 for each, Fig. 4C-D) and diminished both the authentic and induced PS (Fig. 4E-F, n=3 for each). The control-GFP/COS implanted embryo under mitotic arrest condition initially displayed the polonaise movements along the authentic midline axis, however, the vorticity decreased time-dependently, resulting in a defective rotating cell flow (n=3, Fig. 4G-G”’, I-I”’, K-K”’, M-M”’, Fig. S6, and movie S4), consistent with the non-COS cell implanted embryos under aphidicolin-treatment (shown in Fig. 2). In contrast, the Vg1/COS implanted embryos under mitotic arrest condition exhibited the polonaise movements predominantly at the induction point of the secondary axis with disrupted authentic cell flow at the site of the authentic midline axis (n=3, Fig. 4H-H”’, J-J”’, L-L”’, N-N”’, and movie S4). These data suggest that the midline axis induction by the single morphogen brings about the polonaise movements, even when the midline structure (i.e. extending PS) is diminished. Taken together, our data show that Vg1 is capable of inducing and reprogramming the large-scale cell flow during gastrulation even though the midline structure is diminished.

Induced ‘polonaise movements’ persist despite defective PS morphogenesis.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and Vg1. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) BrdU incorporation in control-GFP/COS- and Vg1/COS-implanted embryos under aphidicolin-treatment. White boxes showed enlarged areas. (E, F) In situ hybridizations for Brachyury (Bra, n=3 for each). Black arrows indicate diminished PSs. (G-H’’’) Trace of cell flow path of electroplated epiblast cells by using Imaris tracking analysis. See also Movie S5. White circles, COS cell implanted site. The initiation of the polonaise movements was set as Time 0 (t=0). (I-J’’’) Vector fields, generated from the images of control-GFP/COS-(G-G”’) and Vg1/COS-implanted embryos (H-H’”) under aphidicolin-treatment. (K-L’’’). Streamlines, visualizing averaged cell flows during each time period of the vector fields. (M-N’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period in the vector fields. Note: vorticity is identified in both a full-rotation and curve, particularly in (L-L’’’). Blue, clockwise; red, counter-clockwise rotation. White asterisks, the polonaise movements at the induced axis which are opposite direction to the authentic midline axis. Note: vorticity is identified in both a full-rotation and curve in (M’’’) and (N’’’). See description in the material and method section.

Summary of the patterns of the cell flow and PS morphogenesis.

(A) Cell flow pattern under experimental manipulation of PS morphogenesis.

(B) Authentic PS morphogenesis under disrupting authentic polonaise movements.

(C) Vg1-induced cell flow pattern under defective PS morphogenesis.

(D) Summary of the relationships between polonaise movements, PS morphogenesis, mitosis, and Wnt/PCP pathway.

Discussion

Here, we described a previously undefined relationship between the polonaise movements and PS morphogenesis (Fig. 5). Our approaches to molecularly manipulating PS morphogenesis revealed that the initiation and early phase of the polonaise movements occur despite diminished PS formation (Fig. 5A). Conversely, the experimental manipulation of the cell flow demonstrated that the authentic PS extends regardless of aberrant flow patterns (Fig. 5B). Further, a single axis-inducing morphogen, Vg1, is capable of inducing the bilateral rotating cell flow even under mitotic arrest, when the ectopic PS is reduced (Fig. 5B, C). Our data support a model in which the PS population is required for maintaining the polonaise movements, as seen in the ΔDEP embryos, while the polonaise movements are not necessarily responsible for PS morphogenesis (Fig. 5D).

A similar mitotic arrest has been described as simultaneously diminishing PS extension and disrupting the polonaise movements (19, 20), but it was not determined whether a lack of PS extension and/or diminished cell population of the PS at the midline led to defective polonaise movements. In this study, our live-imaging system combined with the quantitative analyses, molecular tools for manipulating PS formation, and the gene expression analyses allowed us to clarify the relationships among PS morphogenesis, the polonaise movements, mitosis, and the Wnt/PCP pathway (Fig. 1, Fig. 2, and Fig. 5A, C). Our data demonstrated that the shape of the PS as a midline structure is not necessarily responsible for initiation of the polonaise movements while the population of the PS cells, expanded by mitosis, is required for sustained polonaise movements (Fig. 1 and 2). Mesendoderm specification is required for large-scale cell flows associated with later stage ingression and intercalation at the PS during gastrulation (27, 50). Here, we have described the distinctive set of cell flows that precede gastrulation in the chick embryo, and the relationship between early and late cell flow has yet to be examined.

The Wnt/PCP pathway has been assumed to be coupled with the polonaise movements, as an evolutionarily conserved mechanism (21, 51). In contrast, another study showed that misexpression of dnWnt11 or dnDsh-ΔPDZ [suppressing both the Wnt canonical and non-canonical pathway, including the Wnt/PCP (43)] maintained the bilateral counter-rotating cell flow along a defective PS (52). Thus, the role of the Wnt/PCP signaling pathway for the polonaise movements has remained unresolved. In this study, our loss-of-function technique, introducing ΔDEP to PS cells, demonstrated that the polonaise movements were preserved along the deformed PS (Fig. 1, and movie S1). Our electroporation-mediated gene-transfection allowed us to restrict gene transduction in area and number of cells (53). Despite the technical limitations, the suppression level of the Wnt/PCP pathway through the ΔDEP-misexpression was sufficient to lead to the PS deformation (Fig. 1 and Fig. S1). The Wnt/PCP pathway potentially works both cell-autonomously and non-cell autonomously (45, 54). In the PS and non-PS epiblast cells, the disruption of the Pk1-localization was identified not only in the ΔDEP-misexpressing cells, but also in neighbors; however, this study is not designed to provide insight into questions of cell autonomy. Further, the Wnt/PCP pathway is a major regulator of cell intercalation (51, 55).

Recent studies have shown that cell intercalation plays an important role in PS morphogenesis and the initiation of the polonaise movements (20, 21, 27-29). Our data demonstrates that suppression of the Wnt/PCP pathway preserves the initiation and maintenance of the polonaise movements, but the lateral positioning of the rotations is enlarged when the PS is deformed (Fig. 1 and Fig. S1). These results suggest that the Wnt/PCP pathway may not be associated with directional cell intercalation for driving the polonaise movements (Fig. 5C).

To investigate the role of the polonaise movements for PS morphogenesis, a technical challenge/limitation is how to stop or disrupt the cell flow while preserving the cellular process intact. For example, previous loss-of-function studies, using knockdown techniques (such as siRNA and inhibitor treatments), identified the necessity of myosins, including pMyosin and actomyosin, for both the large-scale cell flow and PS morphogenesis, through regulation of tissue tension/biomechanical force (27, 28). Myosins are fundamental cellular components and involved in several cell biological processes, such as tissue tension, cell division, and intracellular transport (56). Therefore, it remains a possibility that myosin-related mechanisms could be involved in the cell flow and PS morphogenesis. Recent studies using injury and pressure models further demonstrated that the cell flow pattern can be altered by a physical cut to an embryonic disc or removal of a large portion of an embryo disc through change of the tissue tension or ablation (20, 57). While these models suggest that the cell flow pattern is dependent on an intact embryonic architecture, it remains unclear as to involvement of other elements such as signaling pathways and gene expression, which can be also molecularly and cell-biologically changed as a result of surgeries. Thus, we developed an experimentally manipulation of the cell flow system, using a Vg1-induced two-midline axis model, disturbing the stereotypical cell flow pattern while maintaining embryonic development (Fig. 3 and Fig. 5B).

Our two-midline axis model provided an important insight into the axis-inducing morphogen for induction and patterning of the polonaise movements (Fig. 3, Fig. 4, Fig. S4-S7, and movie S3, S4). Our data demonstrated that the ectopic Vg1, mimicking the inducing activity of the posterior marginal zone, generated the bilateral rotating cell flow aligned to the induced midline axis but disrupted the cell flows at the authentic midline axis (Fig. 3 and Fig. 4). Despite the disturbed cell flow, PS extension was preserved along both authentic and induced midline axes (Fig. 4). While this model ultimately disturbed the stereotypical cell flow pattern at the authentic midline axis, cell flow was initiated there. Therefore, our study does not rule out potential roles of the cell flow in the initiating steps of PS extension.

It remains unclear how the induced polonaise movements along with the secondary axis overcame the authentic cell flows despite mitotic arrest (Fig. 4 and Fig. 5). As for axis-induction by other axis-inducing morphogens (58), Vg1 may work dose-dependently for induction and patterning of the polonaise movements, and ectopic application may have over-ridden endogenous levels (Fig. 3 and Fig. 4). Further studies will clarify the relationship between the large-scale cell flows and the concentration of the morphogen; this will add new insight to the regulation mechanisms that connect axis-induction and the cell flow patterning.

This study further suggests evolutional distinctions underlying cell biological mechanisms of the large-scale cell flow and midline morphogenesis, between amniotes and non-amniotes. The first distinction is the necessity of the Wnt/PCP pathway to the large-scale cell flow. Whereas the Wnt/PCP pathway is coupled with both the large-scale cell flow and midline morphogenesis in non-amniotes (55), our data revealed that the signaling pathway is required for proper PS morphogenesis, but contributes little to the polonaise movements (Fig. 1). The second distinction is mitosis-dependency/independency comparisons between model systems of gastrulation (Fig. 2). Non-amniotes, in particular amphibians, utilize a blastopore-mediated gastrulation and undergo notochord formation with the large-scale cell flow, all independent of mitosis (10). In amniotes, mitosis is required for PS morphogenesis, and expanding the mass of the PS cells contributes to the maintenance of the polonaise movements (Fig. 2). Thus, the large-scale cell flow and midline morphogenesis in amniote gastrulation is not tightly coupled, which may be evolutionarily distinct from non-amniote gastrulation.

In summary, our study illustrates the relationships between midline morphogenesis, the large-scale cell flow, and an axis-inducing factor in amniote gastrulation. These findings will add another layer of our understanding to how the bilateral body plan is initiated and patterned along the midline axis.

Supporting information

eLife Assessment

Large scale cell movements occur during gastrulation in vertebrate embryos but their role in this major morphogenetic transition in formation of the body plan is poorly understood. Using the chick embryo model system, this study makes important advances using elegant methods to show that extension of the primitive streak during gastrulation, occurring through cell proliferation, polarisation and intercalation, and large-scale polonaise cell movements, can be uncoupled. Although the driving mechanism and precise role of these movements remains a mystery, the study provides convincing evidence for the uncoupling through independent approaches, the most creative of which are the effects shown after induction of a supernumerary primitive streak.

Reviewer #1 (Public Review):

In chicken embryos, the counter-rotating migration of epiblast cells on both sides of the forming primitive streak (PS), a process referred to as polonaise movements, has attracted longstanding interest as a paradigm of morphogenetic cell movements. However, the association between these cell movements and PS development is still controversial. This study investigated PS development and polonaise movements separately at their initial stage, showing that both could be uncoupled (at least at the initial phase), being activated via Vg1 signaling.

Strengths of this study

Polonaise movements, i.e., the circular cell migration of epiblast cells on both sides of the forming PS in avian embryos, have been the subject of research through live imaging and promoted the development of new tools to analyze quantitatively such movements. However, conclusions from previous studies remain controversial, at least partly due to the nature of perturbations to PS development and polonaise movements.

This study performed the challenging technique of electroporation to successfully mark and manipulate Wnt/PCP pathways in unincubated chicken embryo cells at the initiation phase of these two processes. In addition, the authors separately altered PS development and polonaise movements: PS development was perturbed by inhibiting either the Wnt/PCP pathway or DNA synthesis using aphidicolin, while polonaise movements were modified by the development of a second PS after engrafting Vg1-expressing COS cells located at the opposite end of the blastoderm. The study concluded that Vg1 elicits both PS development and polonaise movements, which occur in a parallel and are not inter-dependent.

To support these conclusions, particle image velocimetry (PIV) of cell trajectories captured by live imaging was performed. These tools delineated visually appealing cell movements and gave rise to vorticity profiles, adding more value to this study.

Weaknesses of this study

Engrafted Vg1-expressing COS cells located at the anterior end of the blastoderm elicited both the development of a second PS and marked bilateral polonaise movements while perturbing these movements along the original PS. How do polonaise movements along the second PS dominate over those along the normal PS? The authors suggested a model in which Vg1 acts in a graded or dose-dependent manner since engrafted COS cells over-expressed Vg1. This model can be tested by reducing the mass of engrafted COS cells. Although the authors propose performing this analysis in further investigations, it would be preferable to incorporate into this study for better consistency.

The authors claim that chicken embryo development is representative of "amniotes," but it does not hold for all groups. Avian and mammal species are exceptional among amniotes in the sense they develop a PS (e.g., Coolen et al. 2008). Moreover, in certain mammalian embryos like mouse embryos, cells laterally to the PS do not move much (Williams et al. 2012). The authors should avoid the generalization that chicken embryos unequivocally represent amniotes as opposed to the observed in non-amniote embryos. The observations in chicken embryos as they stand are significant enough.

References:

Coolen M, et al. (2008). Molecular characterization of the gastrula in the turtle Emys orbicularis: an evolutionary perspective on gastrulation. PLoS One. 3(7):e2676. doi: 10.1371/journal.pone.0002676

Williams M, et al. (2012). Mouse primitive streak forms in situ by initiation of epithelial to mesenchymal transition without migration of a cell population. Dev Dyn. 241(2):270-283. doi: 10.1002/dvdy.23711

Reviewer #2 (Public Review):

Summary:

The authors are interested in large-scale cell flow during gastrulation and in particular in the polonaise movement. This movement corresponds to a bilateral vortex-like counter-rotating cell flow and transport the mesendodermal cells allowing ingression of cells through the primitive streak and ultimately the formation of the mesoderm and endoderm. The authors specifically wanted to investigate the coupling of the polonaise movement and primitive streak to understand whether the polonaise movement is a consequence of the formation of the primitive streak or the other way around. They propose a model where the primitive streak elongation is not required for the cell flow but rather for its maintenance and that robust cell flow is not required for primitive streak extension.

Strengths:

Overall, the manuscript is well written with clear experimental designs. The authors have used live imaging and cell flow analysis in different conditions, where either the formation of the primitive streak or the cell flow was perturbed.

Their live imaging and PIV-based analyses convincingly support their conclusions that primitive streak deformation or mitotic arrest do not impact the initiation of the polonaise movement but rather the location or maintenance of these rotations. They additionally showed that disruption of the polonaise movement in the authentic primitive streak by elegant addition of an ectopic primitive streak does not impact the original primitive streak elongation.

Weaknesses:

- When using the delta-DEP-GFP construct, the authors showed that they can manipulate the shape of the primitive streak without affecting the identity and number of primitive streak cells. It is not clear however how this can affect the shape, volume or adhesion of the cells. Some mechanistic insights would strengthen the paper.

- Overall, frequencies of observation are missing for a better view of the phenomenon. For example, do Vg1/Cos cells always disrupt the flow at the authentic primitive streak? Can replicate vector fields be integrated to reflect quantification?

- Since myosin cables have been shown to be instrumental for the polonaise movement, it would be interesting to better investigate how the manipulations by the delta-DEP-GFP construct, or Vg1/Cos affect the myosin cables (as shown in preliminary form for the aphidicolin-treated embryos).

Author Response

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

In chicken embryos, the counter-rotating migration of epiblast cells on both sides of the forming primitive streak (PS), a process referred to as polonaise movements, has attracted longstanding interest as a paradigm of morphogenetic cell movements. However, the association between these cell movements and PS development is still controversial. This study investigated PS development and polonaise movements separately at their initial stage, showing that both could be uncoupled (at least at the initial phase), being activated via Vg1 signaling.

Strengths of this study

Polonaise movements, i.e., the circular cell migration of epiblast cells on both sides of the forming PS in avian embryos, have been the subject of research through live imaging and promoted the development of new tools to analyze quantitatively such movements. However, conclusions from previous studies remain controversial, at least partly due to the nature of perturbations to PS development and polonaise movements.

This study performed the challenging technique of electroporation to successfully mark and manipulate Wnt/PCP pathways in unincubated chicken embryo cells at the initiation phase of these two processes. In addition, the authors separately altered PS development and polonaise movements: PS development was perturbed by inhibiting either the Wnt/PCP pathway or DNA synthesis using aphidicolin, while polonaise movements were modified by the development of a second PS after engrafting Vg1-expressing COS cells located at the opposite end of the blastoderm. The study concluded that Vg1 elicits both PS development and polonaise movements, which occur in a parallel and are not inter-dependent.

To support these conclusions, particle image velocimetry (PIV) of cell trajectories captured by live imaging was performed. These tools delineated visually appealing cell movements and gave rise to vorticity profiles, adding more value to this study.

Weaknesses of this study

Engrafted Vg1-expressing COS cells located at the anterior end of the blastoderm elicited both the development of a second PS and marked bilateral polonaise movements while perturbing these movements along the original PS. How do polonaise movements along the second PS dominate over those along the normal PS? The authors suggested a model in which Vg1 acts in a graded or dose-dependent manner since engrafted COS cells over-expressed Vg1. This model can be tested by reducing the mass of engrafted COS cells. Although the authors propose performing this analysis in further investigations, it would be preferable to incorporate into this study for better consistency.

We would like to express our gratitude to the editors and the reviewers for finding the valuable significances of our study and for giving thoughtful suggestions. We agree that it would be a logical next step to identify the driving mechanism(s) of the polonaise movements, although this is beyond the scope of the current study. Rather, it is the focus of ongoing studies, in which we are investigating how Vg1 works in this concentration context and resulting dose-dependent effect on downstream gene expression, in order to provide a comprehensive understanding of this interesting dual role of Vg1. The relationship between the intensity of Vg1 signaling and the polonaise movements can be tested by modifying the size of the Vg1/COS, as the reviewer pointed out.

The authors claim that chicken embryo development is representative of "amniotes," but it does not hold for all groups. Avian and mammal species are exceptional among amniotes in the sense they develop a PS (e.g., Coolen et al. 2008). Moreover, in certain mammalian embryos like mouse embryos, cells laterally to the PS do not move much (Williams et al. 2012). The authors should avoid the generalization that chicken embryos unequivocally represent amniotes as opposed to the observed in non-amniote embryos. The observations in chicken embryos as they stand are significant enough.

References:

Coolen M, et al. (2008). Molecular characterization of the gastrula in the turtle Emys orbicularis: an evolutionary perspective on gastrulation. PLoS One. 3(7):e2676. doi: 10.1371/journal.pone.0002676

Williams M, et al. (2012). Mouse primitive streak forms in situ by initiation of epithelial to mesenchymal transition without migration of a cell population. Dev Dyn. 241(2):270-283. doi: 10.1002/dvdy.23711

We modified the following sentences to the summary and introduction of the revised version as below:

In Summary:

(p.1, Lines 9-11.) “Large-scale cell flow characterizes gastrulation in animal development. In amniote gastrulation, particularly in avian gastrula, a bilateral vortex-like counter-rotating cell flow, called ‘polonaise movements’, appears along the midline.”

In Introduction:

(p.2, Lines 43-46.) “In amniotes, particularly in avian gastrula (i.e. embryonic disc), a bilateral vortex-like counter-rotating cell flow, termed ‘polonaise movements’, occurs within the epiblast along the midline axis, prior to and during primitive streak (PS) formation.”

Reviewer #2 (Public Review):

Summary:

The authors are interested in large-scale cell flow during gastrulation and in particular in the polonaise movement. This movement corresponds to a bilateral vortex-like counter-rotating cell flow and transport the mesendodermal cells allowing ingression of cells through the primitive streak and ultimately the formation of the mesoderm and endoderm. The authors specifically wanted to investigate the coupling of the polonaise movement and primitive streak to understand whether the polonaise movement is a consequence of the formation of the primitive streak or the other way around. They propose a model where the primitive streak elongation is not required for the cell flow but rather for its maintenance and that robust cell flow is not required for primitive streak extension.

Strengths:

Overall, the manuscript is well written with clear experimental designs. The authors have used live imaging and cell flow analysis in different conditions, where either the formation of the primitive streak or the cell flow was perturbed.

Their live imaging and PIV-based analyses convincingly support their conclusions that primitive streak deformation or mitotic arrest do not impact the initiation of the polonaise movement but rather the location or maintenance of these rotations. They additionally showed that disruption of the polonaise movement in the authentic primitive streak by elegant addition of an ectopic primitive streak does not impact the original primitive streak elongation.

Weaknesses:

When using the delta-DEP-GFP construct, the authors showed that they can manipulate the shape of the primitive streak without affecting the identity and number of primitive streak cells. It is not clear however how this can affect the shape, volume or adhesion of the cells. Some mechanistic insights would strengthen the paper.

We appreciate the reviewer’s invaluable feedback. We agree that it would be informative to know how the ΔDEP-GFP construct led to PS deformation. This approach has been previously introduced by Voiculescu et al., (2007) to demonstrate an involvement of the Dsh(DEP) in PS shape regulation as described in text (please see pp4-5, lines 91-94 in Results and p13, lines 279-281 in Discussion). The previous study suggested that the Wnt/PCP pathway through Dsh(DEP) is a major regulator of cell intercalation, which plays an important role in PS morphogenesis (Voiculescu et al., 2007).

Overall, frequencies of observation are missing for a better view of the phenomenon. For example, do Vg1/Cos cells always disrupt the flow at the authentic primitive streak? Can replicate vector fields be integrated to reflect quantification?

We agree and have added the numbers of embryos examined. In our experimental system, the Vg1/COS-implanted embryos always exhibited that the original polonaise movements along the authentic PS were always disrupted by the induced polonaise movements (n=4/4 embryos). The replicated vector fields were integrated to the Streamline and Vorticity plots (please see Fig. 1-4, Fig. S1, S4-7).

Since myosin cables have been shown to be instrumental for the polonaise movement, it would be interesting to better investigate how the manipulations by the delta-DEP-GFP construct, or Vg1/Cos affect the myosin cables (as shown in preliminary form for the aphidicolin-treated embryos).

We agree that investigations of cytoskeletons and motor proteins would provide deeper understandings as to how the ΔDEP-GFP construct and perhaps Wnt/PCP components work in PS formation and morphogenesis. We plan to examine, as a future study, the patterns of the myosin cables in the ΔDEP-GFP-misexpressing or Vg1/COS-implanted embryos to get better understanding the mechanism(s) of the polonaise movements as the reviewer pointed out.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

The authors named the dominant-negative Dsh lacking DEP [dnDsh(deltaDEP)]-fused GFP as deltaDEP-GFP, presumably to distinguish it from the construct dnDsh-deltaPDZ previously reported. However, the prefix "dnDsh" conveys the critical function in the present study. The reviewer recommends spelling out dnDsh(deltaDEP)-GFP to clarify to readers which signal was manipulated.

We agree that it is necessary to distinguish our construct used in this study from the dnDsh-deltaPDZ construct. We have, therefore, clarified the abbreviation in the main text as follows (please see pp 4-5, lines 91-97): ‘The DEP domain of Dishevelled (Dsh; a transducer protein of Wnt signaling) is responsible for the non-canonical Wnt/PCP pathway (43, 44), and misexpression of dominant-negative Dsh lacking DEP [dnDsh(ΔDEP)] leads to deformation of the midline structures, including the PS (21). Further, the Wnt/PCP pathway is involved in cellular polarity and migration, while the canonical Wnt pathway regulates cell proliferation (45). We refer the dnDsh(ΔDEP)-GFP construct that we generated, as ΔDEP-GFP, and tested its ability to alter cellular polarity, resulting in PS deformation’.

The authors described the "Vg1 plasmid DNA" as a gift from Claudio D. Stern and Jane Dodd. However, they should indicate the vector backbone, especially whether the vector carries the SV40 ori sequence. Ori-containing plasmids multiply after transfection as COS cells express the SV40T antigen, leading to protein overexpression.

We added the name of the plasmid ‘pMT23-Vg1-myc-GDF1’ to the ‘Material and methods’ section (please see p25, line 574). pMT23 expression vector is a derivative of pMT21 (Hume and Dodd, 1993) and contains SV40 ori (Wong et al., 1985).

Reviewer #2 (Recommendations For The Authors):

Most of the comments are indicated in the public review.

There are additionally minor modifications that would help readers interpret the figures. In Figure S1B and D, it is not clear to the reader what the asterisks indicate.

We added the sentence ‘The white asterisks indicate GFP-expressing cells.’ to the figure legend of the Fig. S1 B and D (please see p34, line 874).

Version 2

Abstract

Summary

Large-scale cell flow characterizes gastrulation in animal development. In amniote gastrulation, particularly in avian gastrula, a bilateral vortex-like counter-rotating cell flow, called ‘polonaise movements’, appears along the midline. Here, through experimental manipulations, we addressed relationships between the polonaise movements and morphogenesis of the primitive streak, the earliest midline structure in amniotes. Suppression of the Wnt/planar cell polarity (PCP) signaling pathway maintains the polonaise movements along a deformed primitive streak. Mitotic arrest leads to diminished extension and development of the primitive streak and maintains the early phase of the polonaise movements. Ectopically induced Vg1, an axis-inducing morphogen, generates the polonaise movements, aligned to the induced midline, but disturbs the stereotypical cell flow pattern at the authentic midline. Despite the altered cell flow, induction and extension of the primitive streak are preserved along both authentic and induced midlines. Finally, we show that ectopic axis-inducing morphogen, Vg1, is capable of initiating the polonaise movements without concomitant PS extension under mitotic arrest conditions. These results are consistent with a model wherein primitive streak morphogenesis is required for the maintenance of the polonaise movements, but the polonaise movements are not necessarily responsible for primitive streak morphogenesis. Our data describe a previously undefined relationship between the large-scale cell flow and midline morphogenesis in gastrulation.

Introduction

Large-scale cell flow during gastrulation is an evolutionarily conserved biological phenomenon in embryogenesis (1–3). During this period, animals with bilateral symmetry, called bilaterians, initiate development of three germ layers (i.e. the ectoderm, mesoderm and endoderm) and midline structures (e.g. the notochord and primitive streak) along the midline axis (4–6). The large-scale cell flow has been linked to early embryonic morphogenesis, including axial structures, through rearrangement of cells and/or transportation of signaling molecules (2, 3). Coupling of the large-scale cell flow and body axis morphogenesis has been extensively studied in invertebrate bilaterians, particularly insect models (7–9). However, in vertebrates, our understanding of whether and how the large-scale cell flow and midline morphogenesis are coupled is limited.

Vertebrates are classified into two groups: non-amniotes (e.g. fish and amphibians) and amniotes (e.g. birds and mammals) (1). In non-amniotes, particularly amphibians, the large-scale cell flow is coupled with formation of the notochord through Wnt/planar cell polarity (PCP) pathway-regulated convergent extension, which does not depend on mitosis (10, 11). In amniotes, particularly in avian gastrula (i.e. embryonic disc), a bilateral vortex-like counter-rotating cell flow, termed ‘polonaise movements’, occurs within the epiblast along the midline axis, prior to and during primitive streak (PS) formation (12–14). The PS is an evolutionarily unique embryonic structure in amniotes, which is not only an organizing center for gastrulation but also the earliest midline structure (6, 15–17). During amniote gastrulation, widespread mitosis has been identified throughout the embryo, including the PS (18), and previous studies showed that mitotic arrest leads to morphologically diminishing PS extension (19, 20). Further, the Wnt/PCP pathway also plays an important role for proper PS extension and patterning (21). Since the chick embryo is readily accessible and closely represents the human gastrula (22, 23), much of what is known about the cell flow and PS development has come from studies using this model system (14). The PS is first visible as a cell cluster at the posterior embryonic disc at the initial streak stage [HH2; Hamburger and Hamilton staging (24)] formed through the influence of several axis inducing factors [e.g. Vg1 and Chordin (25, 26)]; it then extends anteriorly along the midline axis (6, 14, 15). During PS development, displacement of cells and/or tissue tension potentially generates biomechanical forces, which lead to cell flows and/or PS morphogenesis (20, 21, 27, 28). Recent studies discuss that cell intercalation and ingression underlying PS formation, leading to the polonaise movements through altering the neighbor tissue tension (21, 28, 29), suggesting that the polonaise movements are likely a consequence of PS morphogenesis (16, 30, 31). Other studies demonstrate that the polonaise movements, generated by the myosin-cable-mediated tissue tension with cell intercalation and ingression, lead to PS morphogenesis (20, 27, 31). Thus, the relationship between the polonaise movements and PS morphogenesis remains controversial (31).

Here, we characterize previously undefined relationships between the polonaise movements and PS morphogenesis. Suppression of the Wnt/PCP pathway generates a wider and shorter PS while maintaining the polonaise movements. Mitotic arrest diminishes PS extension and maintains the early phase of the polonaise movements. Moreover, experimentally manipulating the large-scale cell flow by inducing a secondary midline axis indicates that the authentic PS forms despite altered cell flows. Ultimately, a single axis-inducing morphogen, Vg1, is capable of initiating the polonaise movements despite defective PS morphogenesis. These results suggest that PS morphogenesis is not responsible for the initiation and early phase of the polonaise movements. Further, the polonaise movements are not necessarily responsible for PS morphogenesis. This study suggests that amniote gastrulation may be evolutionarily distinct from non-amniote gastrulation in the mechanisms coupling large-scale cell flow and midline morphogenesis.

Results

Suppression of the Wnt/PCP pathway maintains the bilateral vortex-like counter-rotating cell flow along a deformed PS

Throughout this work, the cell flows were recorded by tagging individual epiblast cells with electroporated fluorescent tracers at the pre-streak stage (32). The tagged embryos were live-imaged until HH3 with a modified New culture system, which we previously described (33, 34). To reconstruct trajectories of the tagged cells and quantitatively analyze the cell flow pattern, the recorded images were processed using tracking analysis tools (35), Particle Image Velocimetry (PIV) (36–40), and subsequent visualization techniques (41, 42).

To address the relationship between the polonaise movements and PS morphogenesis, we experimentally manipulated the shape of the PS while maintaining the population of PS cells (Fig. 1A-B). The DEP domain of Dishevelled (Dsh; a transducer protein of Wnt signaling) is responsible for the non-canonical Wnt/PCP pathway (43, 44), and misexpression of dominant-negative Dsh lacking DEP [dnDsh(ΔDEP)] leads to deformation of the midline structures, including the PS (21). Further, the Wnt/PCP pathway is involved in cellular polarity and migration, while the canonical Wnt pathway regulates cell proliferation (45). We refer the dnDsh(ΔDEP)-GFP construct that we generated, as ΔDEP-GFP, and tested its ability to alter cellular polarity, resulting in PS deformation’ (Fig. 1C-E, Fig. S1A-D). In the control-GFP-introduced PS and epiblast cells, Pk1 (a PCP core component) exhibited a polarized localization at the E-cadherin-based adherens junctions, particularly cell-cell junctions (Fig. S1A-B, n=3). In contrast, this Pk1-localization pattern was not identified in the ΔDEP-GFP-misexpressing PS and epiblast cells (Fig. S1C-D, n=3), implying that the construct efficiently disrupted the cellular polarity within the misexpressing cells. The expression of Brachyury indicated that ΔDEP-GFP-PS cells maintained their identity (Fig. 1E) but displayed a shorter and wider morphogenesis than the control PS (Fig. 1C-E, n≥4 for each). The number of PS cells was not significantly different from the control- and ΔDEP-GFP-PS (control 8.5±0.4×103 cells vs ΔDEP 8.8±1.2×103 cells, p=0.68, n=3 for each, Fig. 1F). These results indicated that ΔDEP-misexpression was capable of suppressing the Wnt/PCP pathway through disrupting the localization pattern of the PCP core component while maintaining the number of the PS cells. Having confirmed the function of the ΔDEP-GFP construct, we applied it as a molecular tool for manipulating the shape of the PS.

‘Polonaise movements’ persist under suppression of Wnt/PCP pathway.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and the Wnt/PCP pathway. A-P and L-R; anterior-posterior and left-right body axes, respectively. (B) Diagram of experimental set up. (C, D) In situ hybridizations for Brachyury (Bra) in control- and ΔDEP-GFP-misexpressing embryos. (E) Overlay of embryos hybridized for Bra (n=4 for each). (F) Cell number in PS [control 8.5±0.4×103 cells vs ΔDEP 8.8±1.2×103 cells, p=0.68, n=3 for each (two-tailed Student’s t test)]. (G-H’’’) Trace of cell flow path of electroplated epiblast cells with Flowtrace. The initiation of the polonaise movements was set as Time 0 (t=0). See also Movie S1. (I-J’’’) Streamlines, visualizing averaged cell flows during each time period. (K-L’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period. Blue, clockwise; red, counter-clockwise rotation.

To address relationship(s) between the polonaise movements and the shape of the PS, we live-imaged the control- and ΔDEP-GFP-embryos from pre-streak stage to HH3 and analyzed the resulting cell flows (Fig. 1G-L”’, and movie S1). In the control-GFP embryo, the cell flow pattern, which was visualized by tracking tools and quantitatively analyzed by PIV with subsequent flow-visualization techniques (41, 42), identified robust polonaise movements (Fig. 1G-G”’, I-I”’, K-K”’, and movie S1, n=5). Surprisingly, these analyses indicated that suppression of the Wnt/PCP pathway maintained the polonaise movements despite PS deformation (n=5, Fig. 1C-L”’ and movie S1). While the topology of the counter-rotating flow pattern were maintained under suppression of the Wnt/PCP pathway (Fig. 1J-J’”), the distance between the left-right rotating cell flows was wider in the ΔDEP embryos than the controls (control 860.2±86.2μm vs ΔDEP 1008.4±183.8μm, p=0.026, n=4 for each, Fig. S1E-I), implying that the shape of the PS and/or the Wnt/PCP signaling-mediated cell intercalation affected the location of these rotations. These data demonstrated that the polonaise movements persisted even though the PS as a midline structure was deformed under the condition of suppressing the Wnt/PCP pathway.

Mitotic arrest diminishes PS formation but initially preserves the bilateral vortex-like rotating cell flow

The above data showed that a deformed midline structure allowed the polonaise movements to occur while the position of the LR rotations in the polonaise movements was affected (Fig. 1 and Fig. S1). However, it remained unclear the relationship between the polonaise movements and the population of the PS cells, particularly expansion of the PS precursor cells in PS morphogenesis (Fig. 2A-B). To address this question, we applied a mitotic arrest approach using aphidicolin (a specific DNA polymerase inhibitor), which has been shown to diminished PS development (19, 20). In chick embryonic disc, active cell proliferation was identified in both the epiblast and PS cells (Fig. 2C)(18). As verified by BrdU incorporation assay, the aphidicolin-treated embryos exhibited little or no mitotic activity (Fig. 2D). Morphogenesis of the PS was monitored by Brachyury (a marker for PS and mesodermal cells) and Sox3 (an epiblast marker). The expression domain and midline extension of cells, expressing Brachyury and negative for Sox3, were considerably reduced in the aphidicolin-treated embryos, compared to the control (Fig. 2E; n≥4 for each). The expression pattern of Brachyury and PS extension followed an aphidicolin dose-dependent diminishment (Fig. S2-D). The Vg1-expressing area was also diminished but detectable in the aphidicolin-treated embryos (Fig. 2E, n=4), suggesting that the axis-inducing morphogen was preserved in the embryonic disc, but expansion of the PS precursor cells was suppressed by mitotic arrest. These results indicate that mitotic arrest, induced by the aphidicolin-treatment, remarkably reduced PS extension and development, consistent with the previous works (19, 20).

Early phase of ‘polonaise movements’ remains under global mitotic arrest.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and mitosis. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) BrdU incorporation in control-sham operated and aphidicolin treated in embryos. White boxes show enlarged areas. (C’, C”) High magnification of boxed areas in PS and non-PS, respectively. (E) In situ hybridizations for Brachyury (Bra), Vg1, and Sox3 in control and aphidicolin-treated embryos. Arrow heads indicate diminished PSs. Black boxes, enlarged area including the diminished PS. (F-G’’’) Trace of cell flow path with Flowtrace analysis (green is trace, yellow indicates endpoint) of fluorescently-tagged epiblast cells. The initiation of the polonaise movements was set as Time 0 (t=0). See also Movie S2. (H-I’’’) Streamlines, visualizing averaged cell flows during each time period. Interpolated vectors are displayed in orange. (J-K’’’) Vorticity plots, displaying averaged measure of the local rotation during each time period. Blue, clockwise; red, counter-clockwise rotation. Note: Since vorticity is calculated for all deviation from set point, slight curves and full rotations receive the same color indication, as seen in K’’’ which is not maintaining a bilateral vortex-like-rotating cell flows. See Materials and Methods section for a full discussion.

To address the relationship between the polonaise movements and the population of the PS cells, we live-imaged the fluorescently tagged embryos under aphidicolin-treatment and analyzed the resulting cell flow (Fig. 2F-K”’ and movie S2). Tracking and PIV-based analyses visualized the cell flow pattern in control and aphidicolin-treated embryos (Fig. 2F-K”’ and movie S2). In control embryos, the polonaise movements occurred from pre-streak until HH3 for more than four hours (n=4, Fig. 2F-F”’ and movie S2). The streamline and vorticity plots illustrated that the polonaise movements became more robust in a time-dependent manner (Fig. 2H-H”’ and J-J”’). Strikingly, in the aphidicolin-treated embryos (n=4), the polonaise movements initiated but were maintained for a shorter period, 2.2±1.2 hours (control 6.5±1.2 hours, p=0.0038, Fig. G-G’”, I-I’”, and movie S2). The streamline and vorticity plots also indicated the initiation and early phase of the polonaise movements in the aphidicolin-treated embryos (Fig. 2J-J”’ and K-K’”). Together, these data demonstrated that the typical pattern of the polonaise movements was induced and initially maintained in the short-term, while there was no or little indication of robust PS formation or extension.

Aphidicolin potentially induces apoptosis (46), which might indirectly lead to the cell flow changes through tissue tension alterations. To test this possibility, we examined apoptosis in the control and aphidicolin-treated embryos (Fig. S3A-B). TUNEL assay showed that the aphidicolin-treatment slightly increased apoptotic rate but there was no significant difference between these embryos and controls (control sham, 10.6±3.5%; aphidicolin, 12.1±3.7%, p=0.20, n=4 for each, Fig. S3B), suggesting that the bilateral rotating cell flow in the aphidicolin-treated embryos was not indirectly induced by apoptosis.

Previous studies have shown that a large-scale ring-like distribution of phospholyrated-myosin (pMyosin) cables among the epiblast cell layer is required for tissue tension related to cell flow (20, 27, 28). We examined whether the pMyosin cables were maintained in the treated embryos over the course of the imaging (Fig. S3C-D”). In the aphidicolin-treated embryos, the large-scale ring-like distribution pattern of the pMyosin cables remained in the epiblast (Fig. S3D-D”), similar to the control embryos (Fig. S3C-C”), implying that the mitotic arrested-embryos maintained tissue tension, consistent with the previous report (20). Thus, these results support a model that mitosis is coupled with PS morphogenesis but is not necessarily required for the initiation and early phase of the polonaise movements.

PS properly extends along the midline despite altering the large-scale cell flow

Given that PS morphogenesis is required for maintenance of the polonaise movements, we next examined the converse of the relationship; that the polonaise movements may be required for PS formation. To test this model, we experimentally reprogrammed the large-scale cell flow (Fig. 3A-B). Ectopically induced Vg1 leads to formation of a Brachyury-positive secondary PS and a secondary midline axis (Fig. 3C-D)(47–49). COS cells expressing either a control-GFP (control-GFP/COS) or Vg1 (Vg1/COS) construct were placed at the anterior marginal zone (AMZ) at pre-streak stage, approximately 180° from the posterior marginal zone where the authentic PS forms, and we analyzed the resulting cell flow during both authentic and ectopically induced PS extension (Fig. 3B). If the bilateral rotating cell flow is required for PS formation, both the authentic and induced PSs should each be accompanied by polonaise movements during the extension. The trajectory analysis by Imaris and PIV-based analyses showed that the control-COS implanted embryos exhibited stereotypical polonaise movements along the authentic midline axis, indicating that the COS cell implantation maintained an authentic cell flow pattern (Fig. 3E-E’”, G-G’”, I-I’”, K, Fig. S4, and movie S3, n=4). In contrast, the analyses of the cell flow in the Vg1/COS implanted embryos illustrated the disruption of the authentic polonaise movements at the extending authentic PS (Fig. 3F-F”’, H-H”’, J-J”’, Fig. S5, and movie S3, n=4). At pre-streak stage, the epiblast cells initially converged to the center of the embryonic disc along both the authentic and the induced axes, and the bilateral rotating cell flows began to appear at both axes (Fig. 3F-F’”, H-H’”, Fig. S5, and movie S3). This cell flow pattern continued for approximately one hour (Fig. 3H’, Fig. S5, and movie S3). Before extension of the authentic PS, the bilateral rotating cell flow along the induced axis became dominant, and eventually overrode the original cell flow (Fig. 3H’-H’”, Fig. S5, and movie S3). Despite global disruption and reversed counter-rotating cell flows, the authentic PS extended properly along the authentic midline (Fig. 3D, F”’, H”’, J”’, K, Fig. S5, and movie S3; n=4). Secondary midline-axis induction by ectopic Vg1 created the polonaise movements around the secondary PS and altered the original cell flow pattern, generating a reprogrammed large-scale cell flow (Fig. 3F-F”’, H-H”’, J-J”’, K, Fig. S6, and movie S3). These results indicate that extension of the authentic PS is preserved even under the defective polonaise movements.

Authentic PS extends under disruption of authentic ‘polonaise movements’.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, using ectopic Vg1-inducing system. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) In situ hybridizations for Brachyury (Bra) in control and ectopic Vg1-induced PS, respectively. (E-F’’’) Trace of cell flow path of electroplated epiblast cells by using Imaris tracking analysis. See also Movie S3. White circles, COS cell implanted site. The initiation of the polonaise movements was set as Time 0 (t=0). (G-H’’’) Streamlines, visualizing averaged cell flows during each time period of the vector fields. (I-J’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period in the vector fields. Blue, clockwise; red, counter-clockwise rotation. Note: vorticity is identified in both a full-rotation and curve, particularly in (J-J’’’). The white asterisks indicate the polonaise movements at the induced PS which are opposite direction to the authentic PS. (K) Summary of the cell flow patterns in control-GFP/COS- and Vg1/COS-implanted embryos, shown in (E-J’”).

Vg1 is capable of inducing the bilateral vortex-like rotating cell flow despite diminished PS extension under mitotic arrest

The above data indicate the lack of a strong relationship between the polonaise movements and PS morphogenesis; however, it remained unclear how the polonaise movements were induced. Given that mitotic arrest maintained both Vg1-expression and the initiation of the polonaise movements (Fig. 2 and movie S2) and that induction of the secondary midline axis by ectopic Vg1 reprograms the epiblast cell flow (Fig. 3, Fig. S4, S5, and movie S3), we tested the possibility that the axis-inducing morphogen, Vg1, would be capable of inducing the polonaise movements in the absence of PS extension (Fig. 4A). We combined the secondary axis induction by AMZ implantation of Vg1/COS cells with aphidicolin-treatment and analyzed the resulting cell flow (Fig. 4B, Fig. S6, S7, and movie S4). The BrdU assay showed that the aphidicolin treatment effectively blocked mitosis in Vg1/COS implanted embryos (control sham 88±9.0% vs control-GFP/COS with aphidicolin 0.2±0.1% vs Vg1/COS with aphidicolin 0.19±0.2%, p= 0.0005×10-3, n=3 for each, Fig. 4C-D) and diminished both the authentic and induced PSs (Fig. 4E-F, n=3 for each). The control-GFP/COS implanted embryo under mitotic arrest condition initially displayed the polonaise movements along the authentic midline axis, however, the vorticity decreased time-dependently, resulting in a defective rotating cell flow (n=3, Fig. 4G-G”’, I-I”’, K-K”’, Fig. S6, and movie S4), consistent with the non-COS cell implanted embryos under aphidicolin-treatment (shown in Fig. 2). In contrast, the Vg1/COS implanted embryos under mitotic arrest condition exhibited the polonaise movements predominantly at the induction point of the secondary axis with disrupted authentic cell flow at the site of the authentic midline axis (n=3, Fig. 4H-H”’, J-J”’, L-L”’, and movie S4). These data suggest that the midline axis induction by the single morphogen brings about the polonaise movements, even when the midline structure (i.e. extending PS) is diminished. Taken together, our data show that Vg1 is capable of inducing and reprogramming the large-scale cell flow during gastrulation even though the midline structure is diminished.

Induced ‘polonaise movements’ persist despite defective PS morphogenesis.

(A) Testing model of relationships between the polonaise movements, primitive streak (PS) morphogenesis, and Vg1. Axes as in Fig. 1. (B) Diagram of experimental set up. (C, D) BrdU incorporation in control-GFP/COS- and Vg1/COS-implanted embryos under aphidicolin-treatment. White boxes showed enlarged areas. (E, F) In situ hybridizations for Brachyury (Bra, n=3 for each). Black arrows indicate diminished PSs. (G-H’’’) Trace of cell flow path of electroplated epiblast cells by using Imaris tracking analysis. See also Movie S5. White circles, COS cell implanted site. The initiation of the polonaise movements was set as Time 0 (t=0). (I-J’’’) Streamlines, visualizing averaged cell flows during each time period of the vector fields. (K-L’’’) Vorticity plots, displaying an averaged measure of the local rotation during each time period. Note: vorticity is identified in both a full-rotation and curve, particularly in (L-L’’’). Blue, clockwise; red, counter-clockwise rotation. White asterisks, the polonaise movements at the induced axis which are opposite direction to the authentic midline axis. Note: vorticity is identified in both a full-rotation and curve in (K’’’) and (L’’’). See description in the material and method section.

Discussion

Here, we described a previously undefined relationship between the polonaise movements and PS morphogenesis (Fig. 5). Our approaches to molecularly manipulating PS morphogenesis revealed that the initiation and early phase of the polonaise movements occur despite diminished PS formation (Fig. 5A). Conversely, the experimental manipulation of the cell flow demonstrated that the authentic PS extends regardless of aberrant flow patterns (Fig. 5B). Further, a single axis-inducing morphogen, Vg1, is capable of inducing the bilateral rotating cell flow even under mitotic arrest, when the ectopic PS is reduced (Fig. 5B, C). Our data support a model in which the PS population is required for maintaining the polonaise movements, as seen in the ΔDEP embryos, while the polonaise movements are not necessarily responsible for PS morphogenesis (Fig. 5D).

Summary of the patterns of the cell flow and PS morphogenesis.

(A) Cell flow pattern under experimental manipulation of PS morphogenesis.

(B) Authentic PS morphogenesis under disrupting authentic polonaise movements.

(C) Vg1-induced cell flow pattern under defective PS morphogenesis.

(D) Summary of the relationships between polonaise movements, PS morphogenesis, mitosis, and Wnt/PCP pathway.

A similar mitotic arrest has been described as simultaneously diminishing PS extension and disrupting the polonaise movements (19, 20), but it was not determined whether a lack of PS extension and/or diminished cell population of the PS at the midline led to defective polonaise movements. In this study, our live-imaging system combined with the quantitative analyses, molecular tools for manipulating PS formation, and the gene expression analyses allowed us to clarify the relationships among PS morphogenesis, the polonaise movements, mitosis, and the Wnt/PCP pathway (Fig. 1, Fig. 2, and Fig. 5A, C). Our data demonstrated that the shape of the PS as a midline structure is not necessarily responsible for initiation of the polonaise movements while the population of the PS cells, expanded by mitosis, is required for sustained polonaise movements (Fig. 1 and 2). Mesendoderm specification is required for large-scale cell flows associated with later stage ingression and intercalation at the PS during gastrulation (27, 50). Here, we have described the distinctive set of cell flows that precede gastrulation in the chick embryo, and the relationship between early and late cell flow has yet to be examined.

The Wnt/PCP pathway has been assumed to be coupled with the polonaise movements, as an evolutionarily conserved mechanism (21, 51). In contrast, another study showed that misexpression of dnWnt11 or dnDsh-ΔPDZ [suppressing both the Wnt canonical and non-canonical pathway, including the Wnt/PCP (43)] maintained the bilateral counter-rotating cell flow along a defective PS (52). Thus, the role of the Wnt/PCP signaling pathway for the polonaise movements has remained unresolved. In this study, our loss-of-function technique, introducing ΔDEP to PS cells, demonstrated that the polonaise movements were preserved along the deformed PS (Fig. 1, and movie S1). Our electroporation-mediated gene-transfection allowed us to restrict gene transduction in area and number of cells (53). Despite the technical limitations, the suppression level of the Wnt/PCP pathway through the ΔDEP-misexpression was sufficient to lead to the PS deformation (Fig. 1 and Fig. S1). The Wnt/PCP pathway potentially works both cell-autonomously and non-cell autonomously (45, 54). In the PS and non-PS epiblast cells, the disruption of the Pk1-localization was identified not only in the ΔDEP-misexpressing cells, but also in neighbors; however, this study is not designed to provide insight into questions of cell autonomy. Further, the Wnt/PCP pathway is a major regulator of cell intercalation (51, 55). Recent studies have shown that cell intercalation plays an important role in PS morphogenesis and the initiation of the polonaise movements (20, 21, 27–29). Our data demonstrates that suppression of the Wnt/PCP pathway preserves the initiation and maintenance of the polonaise movements, but the lateral positioning of the rotations is enlarged when the PS is deformed (Fig. 1 and Fig. S1). These results suggest that the Wnt/PCP pathway may not be associated with directional cell intercalation for driving the polonaise movements (Fig. 5C).

To investigate the role of the polonaise movements for PS morphogenesis, a technical challenge/limitation is how to stop or disrupt the cell flow while preserving the cellular process intact. For example, previous loss-of-function studies, using knockdown techniques (such as siRNA and inhibitor treatments), identified the necessity of myosins, including pMyosin and actomyosin, for both the large-scale cell flow and PS morphogenesis, through regulation of tissue tension/biomechanical force (27, 28). Myosins are fundamental cellular components and involved in several cell biological processes, such as tissue tension, cell division, and intracellular transport (56). Therefore, it remains a possibility that myosin-related mechanisms could be involved in the cell flow and PS morphogenesis. Recent studies using injury and pressure models further demonstrated that the cell flow pattern can be altered by a physical cut to an embryonic disc or removal of a large portion of an embryo disc through change of the tissue tension or ablation (20, 57). While these models suggest that the cell flow pattern is dependent on an intact embryonic architecture, it remains unclear as to involvement of other elements such as signaling pathways and gene expression, which can be also molecularly and cell-biologically changed as a result of surgeries. Thus, we developed an experimentally manipulation of the cell flow system, using a Vg1-induced two-midline axis model, disturbing the stereotypical cell flow pattern while maintaining embryonic development (Fig. 3 and Fig. 5B).

Our two-midline axis model provided an important insight into the axis-inducing morphogen for induction and patterning of the polonaise movements (Fig. 3, Fig. 4, Fig. S4-S7, and movie S3, S4). Our data demonstrated that the ectopic Vg1, mimicking the inducing activity of the posterior marginal zone, generated the bilateral rotating cell flow aligned to the induced midline axis but disrupted the cell flows at the authentic midline axis (Fig. 3 and Fig. 4). Despite the disturbed cell flow, PS extension was preserved along both authentic and induced midline axes (Fig. 4). While this model ultimately disturbed the stereotypical cell flow pattern at the authentic midline axis, cell flow was initiated there. Therefore, our study does not rule out potential roles of the cell flow in the initiating steps of PS extension.

It remains unclear how the induced polonaise movements along with the secondary axis overcame the authentic cell flows despite mitotic arrest (Fig. 4 and Fig. 5). As for axis-induction by other axis-inducing morphogens (58), Vg1 may work dose-dependently for induction and patterning of the polonaise movements, and ectopic application may have over-ridden endogenous levels (Fig. 3 and Fig. 4). Further studies will clarify the relationship between the large-scale cell flows and the concentration of the morphogen; this will add new insight to the regulation mechanisms that connect axis-induction and the cell flow patterning.

This study further suggests evolutional distinctions underlying cell biological mechanisms of the large-scale cell flow and midline morphogenesis, between amniotes and non-amniotes. The first distinction is the necessity of the Wnt/PCP pathway to the large-scale cell flow. Whereas the Wnt/PCP pathway is coupled with both the large-scale cell flow and midline morphogenesis in non-amniotes (55), our data revealed that the signaling pathway is required for proper PS morphogenesis, but contributes little to the polonaise movements (Fig. 1). The second distinction is mitosis-dependency/independency comparisons between model systems of gastrulation (Fig. 2). Non-amniotes, in particular amphibians, utilize a blastopore-mediated gastrulation and undergo notochord formation with the large-scale cell flow, all independent of mitosis (10). In amniotes, mitosis is required for PS morphogenesis, and expanding the mass of the PS cells contributes to the maintenance of the polonaise movements (Fig. 2). Thus, the large-scale cell flow and midline morphogenesis in amniote gastrulation is not tightly coupled, which may be evolutionarily distinct from non-amniote gastrulation.

In summary, our study illustrates the relationships between midline morphogenesis, the large-scale cell flow, and an axis-inducing factor in amniote gastrulation. These findings will add another layer of our understanding to how the bilateral body plan is initiated and patterned along the midline axis.

Supporting information